---
title: "Introduction to Social Network Analysis"
author: "Chanya Punyakumpol"
output: 
  html_document:
    toc: true
    toc_float: true
date: "Last edited `r format(Sys.Date(), '%B %d, %Y')`"
params:
  url: https://www.designoftradeagreements.org/media/filer_public/ab/ee/abee41ef-f5e5-44b6-91db-5e8befe48fe5/desta_list_of_treaties_02_01_dyads.xlsx
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', echo = TRUE)
```

# I. Getting to know the dataset
Welcome to Introduction to Social Network Analysis workshop! This file will introduce to you the concept of social network analysis and its application to international law research. The workshop is specifically designed for law students with some limited backgrounds in R. If you have no knowledge or no background in R, I highly recommend that you take some basic introductions to R, including working with vectors, lists, matrices, data wrangling, and data visualization. Hence, in taking this workshop, I assume that you have some of those basic knowledge of R.

The objectives of this workshop are:

- to familiarize you with the concept of social network in a practical sense by working on a real data set;
- to walk you through all steps necessary to create a network in R, as well as simple network visualizations;
- to introduce basic features of a social network that provides insights into a relationship of interests; 
- to help you learn how to perform basic network analytics such as clustering and community detection; and
- to introduce you to some advanced topics that may be of interest to you.

At the end of the workshop, you should be able to adapt what you have learned to your own legal research.

## A. Setting up the workspace

First, create a new R Project on your computer. Make sure that you know where you save this R project. All of the outputs you have will be saved in this folder. Next, run the following lines of code to download the necessary R packages for this workshop.

``` {r load necessary packages, message = FALSE}
#if any error message occurred, make sure you have installed these packages first before you load them
library(readxl) #to import data
library(dplyr) #for data wrangling
library(knitr) #to show table on RMarkdown
library(ggplot2) #for data visualization
library(igraph) #we will primarily use igraph package in this workshop
```

Second, you will have to import the data, which we will use throughout the workshop, which you can inspect [here](https://www.designoftradeagreements.org/media/filer_public/ab/ee/abee41ef-f5e5-44b6-91db-5e8befe48fe5/desta_list_of_treaties_02_01_dyads.xlsx). The data set is from the [DESTA](https://www.designoftradeagreements.org/) (Design of Trade Agreements) Database, a project that contains comprehensive data on various types of preferential trade agreements (PTAs).

Now, you are ready to import the data to your RStudio workspace:

```{r data, message = FALSE}
excel_file <- tempfile()
download.file(params$url,excel_file, mode = "wb")
pta_data <- read_excel(path = excel_file, sheet = 2)
pta_data$regioncon <- as.factor(pta_data$regioncon)
```

```{r inspect the data}
#inspect the data
str(pta_data) #look at the structure of the data
head(pta_data) #look at the first few rows of the data
```

## B. Exploring the data
As you can see, the data we are dealing with is very large. Exploring data is crucial for quantitative approach to help us learn about the data, as well as to help us figure out what to do next with the data. We do so by trying to find some patterns from this data set first. Based on the data, two variables that could be of interest here are: year and the region to which a PTA belongs. A simple way to do this is to plot the data using ggplot2 package:

```{r data exploration}
pta_data_count <- pta_data %>% group_by(year, regioncon) %>% count(name) #this line counts the frequency of ties between any two countries formed by year and region.
pta_data_count <- pta_data_count %>% arrange(desc(n))
kable(head(pta_data_count), col.names = c("year", "region", "agreement name", "number of ties"),
      caption = "Frequency of Ties Table")
kable(summary(pta_data_count), col.names = c("year", "region", "agreement name", "number of ties"), 
      caption = "A summary of the Frequency of Ties Data") #summary() provide a summary of the data
```
From a summary of data, notice that there are six different categories: Africa, Americas, Asia, Europe, Oceania, and Intercontinental. Most PTAs are intercontinental (382 agreements), while not surprisingly Oceania has only 12 intra-regional agreements. This is due to the number of member countries in the region. 

Next, we are going to look at which countries tend to form multilateral PTAs and which countries tend to form bilateral PTAs:

```{r regional preference}
#filter the new data
pta_data_count <- pta_data_count %>% mutate(bilat = ifelse(n == 1, "bilateral", "multilateral")) #creating a new variable to see if a tie belongs to a bilateral or multilateral treaty
pta_data_count$bilat <- as.factor(pta_data_count$bilat)
region_count <- pta_data_count %>% group_by(regioncon, bilat) %>% count(regioncon)
ggplot(region_count, aes(x= regioncon, y = n, color = bilat, fill = bilat)) +
  geom_col()
```
Out of this simple bar plot, we notice that countries in Africa prefer to have multilateral PTAs rather than bilateral ones, while countries in Americas and Asia clearly prefer to form bilateral PTAs instead of multilateral ones. Europe and Oceania are almost equally split between bilateral multilateral PTAs. Interestingly, the inter-regional PTAs are also ambivalent with regards to bilateral and multiplateral PTAs.

```{r scatter plot}
ggplot(pta_data_count, aes(x = year, y = n, color = regioncon, fill = regioncon)) +
  geom_col() + facet_wrap(~regioncon)
```
Indeed, plotting how many ties formed each year by regions confirm our hypothesis that Africa stands out with the number of ties formed, along with intra-regional PTAs, followed by Europe. Asia, Americas and Oceania prefers to form bilateral agreements.

# II. Creating a network
Once we have preliminary explored the data, one of the directions we can take is to look at the inter-regional network because that is where PTAs have the most impact on inter-regional trade flows. What else could be further explored based on the data exploratory session? This is a food for thought, as well as for you to try social network analysis on your own.

In this section, I will introduce how to create a network in R. There are two types of data structure that you can setup to construct:

1. Edgelist
2. Adjacency Matrix

The edgelist is exactly what this DESTA database has done: the dyadic country relationship as shown in the first two coloumns, `country1`, and `country2`,  the original data set here:

```{r edgelist data structure}
head(pta_data)
```
Although to construct a simple network, you do not need other variables in the data set, it is useful to keep these variables to, as you will see later, set the tie properties in the network.

Alternatively, you can also construct a network from an adjacency matrix as the underlying data structure.

```{r turn the first six rows to an adjacency matrix}
#ignore these two lines first, we will cover these functions later on
#the point is to show an alternative data structure available to construct a network show in the output here
temp_graph <- graph_from_edgelist(cbind(head(pta_data$country1), head(pta_data$country2)), directed = F)
get.adjacency(temp_graph)
```

## A. From the edgelist
The function in `igraph` package, called `graph_from_edgelist()` will create a network from the edgelist data for you: 
```{r edgelist network temp}
cbind(head(pta_data$country1), head(pta_data$country2))
edge_net <- graph_from_edgelist(cbind(head(pta_data$country1), head(pta_data$country2)),
                                directed = F) #you have created the first network from edgelist!

#see the network:
set.seed(123) #set.seed() fixes the configuration of the plot for the sake of reproducibility and comparison in this case
plot.igraph(edge_net) #don't worry about this we will cover this function in the next section
```

## B. From the adjacency matrix
To create a network from an adjacency matrix, call the function `graph_from_adjacency_matrix()` from the `igraph` package.

```{r adjacency matrix network temp}
get.adjacency(edge_net)
matrix_net <- graph_from_adjacency_matrix(get.adjacency(edge_net), mode = "undirected")
set.seed(123) #again, set.seed() here to make sure the plot looks similar
plot.igraph(matrix_net)
```
As you can see, the two networks look exactly the same!

## C. Construct the network of countries by their inter-regional PTAs
Now, let's construct the network of countries by their inter-regional PTAs, using the edgelist data already provided in the DESTA data set. Because the data structure is an edgelist data, we use `graph_from_edgelist()`. 

```{r construct network}
pta_intercon <- pta_data %>% filter(regioncon == "Intercontinental")
pta_intercon_net <- graph_from_edgelist(cbind(pta_intercon$country1, pta_intercon$country2), directed = F) 
```
Now that we have successfully created the network, the next step is to assign the value to the ties for further uses. Useful variables to be assigned for the purpose of this workshop are: `name` and `year`. The function `E()` calls an edge sequence of the network, as identified by the corresponding indices. The function can also be used to assign edge based attributes. We will use this to assign some properties of the edges and the weights of edges. Here, we call the edges of the network, using `E()` followed by `$` and the name of the variable we wish to have, `agt_name` for the PTAs' names, and `year` for the year of these PTAs.

```{r add other edge attributes}
E(pta_intercon_net)$agt_name <- pta_intercon$name
E(pta_intercon_net)$year <- pta_intercon$year
```

Note that there might be duplicated edges between a pair of countries, this is the case when the two countries have formed more than one PTA together. Instead of creating duplicated edges, we can assign a weight to each pair corresponding to the number of PTAs shared by the two countries. Next, using the `simplify()` function from the `igraph` package to simplify the network by removing loops and multiple edges. In doing so, `simplify()` can assign the sum of all duplicated edges to the `weight` of that edge.

```{r simplify network}
#set weighted ties:
E(pta_intercon_net)$weight <- 1
pta_intercon_net <- pta_intercon_net %>% simplify(edge.attr.comb = list(weight = "sum"))
```

It is also important to recognize that once we simplify the edges of the network, we will lose some information about `name` and `year` variables to those edges. The simplified network can be useful for the purpose of looking at the descriptives (to be covered in the next section). However, we the names or the years of PTAs are of importance to your analysis, you may have to revert back to the full network.

# III. Descriptives of the network
After we have successfully created the network, in this section, we will learn to explore some network properties -- the descriptives of the network. First, we will look at how to visualize the network. Thereafter, the basic network descriptions such as densitiy, degree distributions, and transitivity will be covered. Next, we will go over various types of network centralities to identify key players in the network.

## A. Simple visualizations

## B. Basic descriptions of the network

## C. Network Centralities

# IV. Clustering and community detection

# V. More advanced topics
