---
title: "Introduction to Social Network Analysis"
author: "Chanya Punyakumpol"
output: 
  html_document:
    toc: true
    toc_float: true
date: "Last edited `r format(Sys.Date(), '%B %d, %Y')`"
params:
  url: https://www.designoftradeagreements.org/media/filer_public/87/e2/87e266f0-5874-4596-913a-251e99ab6632/desta_list_of_treaties_02_01_dyads.csv
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', echo = TRUE)
```

# I. Getting to know the dataset
Welcome to Introduction to Social Network Analysis workshop! This file will introduce to you the concept of social network analysis and its application to international law research. The workshop is specifically designed for law students with some limited backgrounds about R. If you have no knowledge or no background about R, I highly recommend that you take some basic introductions to R, including working with vectors, lists, matrices, data wrangling, and data visualization. Hence, in taking this workshop, I assume that you have some of those basic knowledge of R.

The objectives of this workshop are:

- to familiarize you with the concept of social network in a practical sense by working on a real data set;
- to walk you through all steps necessary to create a network in R, as well as simple network visualizations;
- to introduce basic features of a social network that provides insights into a relationship of interests; 
- to help you learn how to perform basic network analytics such as clustering and community detection; and
- to introduce you to some advanced topics that may be of interest to you.

At the end of the workshop, you should be able to adapt what you have learned to your own legal research.

## A. Setting up the workspace

First, create a new R Project on your computer. Make sure that you know where you save this R project. All of the outputs you have will be saved in this folder. Next, run the following lines of code to download the necessary R packages for this workshop.

``` {r load necessary packages, message = FALSE}
#if any error message occurred, make sure you have installed these packages first before you load them
library(readr) #to import data
library(dplyr) #for data wrangling
library(ggplot2) #for data visualization
library(igraph) #we will primarily use igraph package in this workshop
```

Second, you will have to import the data, which we will use throughout the workshop, which you can inspect [here](https://www.designoftradeagreements.org/media/filer_public/87/e2/87e266f0-5874-4596-913a-251e99ab6632/desta_list_of_treaties_02_01_dyads.csv). The data set is from the [DESTA](https://www.designoftradeagreements.org/) (Design of Trade Agreements) Database, a project that contains comprehensive data on various types of preferential trade agreements (PTAs).

Now, you are ready to import the data to your RStudio workspace:

```{r data, message = FALSE}
pta_data <- read_csv(params$url) #the url is already predefined as a global parameter here.
```

```{r inspect the data}
#inspect the data
str(pta_data)
head(pta_data)
```

## B. Simplifying the data
As you can see, there are so many




# II. Creating a network

## A. From the edgelist

## B. From the adjacency matrix

# III. Descriptives of the network

## A. Simple visualizations

## B. Basic descriptions of the network

## C. Network Centralities

# IV. Clustering and community detection

# V. More advanced topics
